---
title: "A Comprehensive Analysis of Trends, Causes, and Recommendations for Toronto's streetcar's delay"
author: YITONG WANG
thanks: "Code and data are available at: https://github.com/demainwang/streetcar_delay_analysis"
date: today
date-format: long
abstract: "This paper analyze the delay patern of streetcar in Toronto in 2023 years based on the data from opendatatoronto. The result of analysis figure out that most streetcars' delays are concentrate in peak hour and occurs in specific intersection which located at prosperous square. Overall delays are short and within 10 minutes, extreme cases occasionaly happen. The main cause of delays are operational issue and mechanical problems. Delay patter and related incidents are siginificated insights to improve the efficiency of public transportation, and it is highly related with citizen's life quality, the finding provides an instructive direction which public transit should consider."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(knitr)
```

```{r}
#| include: false
#| warning: false
#| message: false
# Read data
cleaned_data <- read_csv('cleaned_data.csv')
```

# Introduction

In the metropolis like Toronto, public transportation is critical and
essential for citizens' mobility, the streetcar runs by TTC is one
common way that people use in their life. Public transportation
originally aims to provide a more convenient way for people's mobility
and reduce the traffic congestion, but the frequent delays can cause an
significant disruption for living quality and business.

It is important to understand the pattern of delays, which can
potentially improve the public transportation system. The reason of
delays can be widely vary with time, location and incident. The data
analysis notice that delay of streetcar has a high possibility occur
during rush hour in some prosperous region such as dunds square. The
statistic analysis show that overall delay's time within 1 hour, and
concentrate in 10 minutes, some extreme cases also happen but with a
less possibility. Due to the construction operating route of streetcar,
most lines are east-west orientation, most delays occure in east-west
bound.

The paper focus on the 2023 streetcar's delay data obtained from
opendatatoronto (@citeopendatatoronto), based on R (@citeR) to do the
data computing, and use package 'tidyverse' (@rohan) to create
visualization and plotting, especially 'ggplot2' (@citeggplot2), 
most plots are created through this package. The usage of 'knitr' (@citeknitr) is for
establishing the summary table. We also use 'hms' (@citehms) package to ensure data about time
work correctlt. The paper aims to figure out the delay's
pattern, and identify the variables contribute to delays of streetcar,
such as location and time period, especially which incidents mainly
affect the normal operation of streetcar. The rest of paper will discuss
the source data and introduce the variables in @sec-data, the following part
is visualization which contains some bar chats to analyze the frequency of
streetcar's delay pattern, and summary table to conclude the paper.

\newpage

# Data {#sec-data}

## Data bacground and measurements

The data we use in this paper comes from opendatatoronto (@citeopendatatoronto), the dataset
'TTC Streetcar Delay Data' provides detailed information on
streetcar delays in Toronto in 2023, which published by Toronto Transit Commission.
This is the official data of the recording of streetcar's delay, all data comes from TTC's original internal records, so the likelihood of bias and errors is very small.
The raw data contains the following variables, 'Date' for date and 'Time' for exact time
when delay happens, in form "HH:MM", the 'Line' for route, the 'Location' for location where delay occurred, the 'Incident' about reason for the delay, 'Min Delay' the duration of the delay. 'Min Gap' for the time of streetcar ahead of the following streetcar. 

## Data Cleaning
We mainly focus on the pattern of occurred time, duration, location and direction of
streetcar to figure out the pattern of occurrence of delay, therefore we do not need any information about route number and vehicle number, and 'Min Gap' is not our focus point.
Since the original document is xlsx, the process of data cleaning as
following steps:

1. The original data mostly looks good, but some values
do not present correctly. We firstly use trimws() to delete any posiible
whitespace nearby the data, then use as.numeric() to make sure data
store in 'Min Gap' column as numeric value. 

2. During these process,
some blank or abnormal data will become to N/A value, we simply use
na.omit() to ignore any rows contain N/A values.

3. We use sum to
glimpse the delay values interval, and figure out that over 99 percents
data within one our, some extreme case may affect the the eprformance of
overall data, so we filter the 'Delay Min" within one hour, which also
omit some 0 minutes delay.

## Ethical and social consideration

From the ethical perspective, this dataset does not contain any personal
information, so there are fewer privacy issues. However, we need to
consider the social context that drives data collection. Toronto's
public transportation system, especially streetcar services, has always
been a concern due to its reliability issues. The collection of this
data aims to improve transparency and provide a foundation for data
based policy making to improve the efficiency of public transportation.

\newpage

## Visualization

```{r}
#| eveal: true
#| label: fig-1
#| fig-cap: "Distribution of streetcar's delay time frequency"
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 1

ggplot(cleaned_data, aes(x = `Min Delay`)) +
  geom_histogram(binwidth = 5, fill = "orange", color = "black") +
  labs(title = "Frequency of Streetcar Delay Duration",
       x = "Delay Duration (Mins)", y = "Frequency") +
  theme_minimal()
```

@fig-1 shows a clear right skewed distribution of the frequency of delay
time period, most common duration is around 10 minutes, which occupy 60
percents of total delay times. It presents in most situation the
streetcar just typically delay for short time, and does not make a
serious disruption for citizen's convenience.

\clearpage

```{r}
#| eval: true
#| label: fig-2
#| fig-cap: "Frequecy of incidents and incidents causing delay time"
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2

incidents <- cleaned_data %>%
  count(Incident, sort = TRUE)
ggplot(incidents, aes(x = reorder(Incident, n), y = n)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Frequency of Streetcar Incidents",
       x = "Incident Type", y = "Frequency of Incident") +
  theme_minimal() +
  coord_flip()

ggplot(cleaned_data, aes(x = Incident, y = `Min Delay`)) +
  geom_boxplot(fill = "orange", color = "black") +
  labs(title = "Delay Time by Vary Incidents",
       x = "Incident", y = "Delay Time (Mins)") +
  theme_minimal() +
  coord_flip()
```

@fig-2-1 The bar plots presents operational issue and mechnical problem
are the top 2 main cause of delay, and the responsibility for this 2
factors belong to TTC, it is obvious TTC company should improve their
internal training and have a better inspection before the operation of
streetcar.

@fig-2-2 Fortunately, the box plot shows these 2 main factors ofdelay do
not result in a hard disruption of citizen's transportation, they
usually lead to a 10 minutes delay.

\clearpage

```{r}
#| eveal: true
#| label: fig-3
#| fig-cap: "Trend of delay times by month"
#| echo: false
#| warning: false
#| message: false

monthly_delay <- cleaned_data %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Month) %>%
  summarize(Delay_Count = n())

ggplot(monthly_delay, aes(x = Month, y = Delay_Count)) +
  geom_line(color = "black", linewidth = 1) +
  geom_point(color = "red", size = 1) +
  labs(title = "Time Series of Streetcar Delays Through the Year",
       x = "Month", y = "Times of Delays") +
  theme_minimal()
```

@fig-3 The line plot shows a increase trend of the frequency of delays,
before september the monthly delay has fluctuation between 750 to 900,
and has a significant increase from september to october, then keep
going up. The incrase trend shows the delay frequency becoming
increasingly serve alone time.

\clearpage

```{r}
#| eval: true
#| label: fig-4
#| fig-cap: "Trend of delay times by month"
#| echo: false
#| warning: false
#| message: false

summary_table <- cleaned_data %>%
  group_by(Location) %>%
  summarise(Total_Delays = n()) %>%  
  arrange(desc(Total_Delays)) %>%
  slice(1:15)
kable(summary_table, caption = "Summary of Streetcar Delays by Location",
      col.names = c("Location", "Total Delays"))
```

@fig-4 This table summary the total delays times by different location,
and most of the top locations are busy intersection nearby the business
district, whcih has a higher demand for public trnsportation, the
TTCcompany should consider this situation and find the solution to
reduce delay.

\clearpage

```{r}
#| eval: true
#| label: fig-5
#| fig-cap: "Monthly summary"
#| echo: false
#| warning: false
#| message: false

cleaned_data <- cleaned_data %>%
  mutate(Date = as.Date(Date, format = "%Y-%m-%d"),
         Month = format(Date, "%Y-%m"))

# Summarize data by Month
summary_table <- cleaned_data %>%
  group_by(Month) %>%
  summarise(
    Average_Delay = mean(`Min Delay`),  
    Total_Delays = n(),                           
    Max_Delay = max(`Min Delay`),                 
    Most_Frequent_Location = Location[which.max(table(Location))],  
    Most_Frequent_Bound = Bound[which.max(table(Bound))]
  )

# Create the table with kable()
kable(
  summary_table, 
  caption = "Monthly Summary of Streetcar Delays",
  col.names = c("Month", "Average Delay (Mins)", "Total Delay Times", "Max Delay (Mins)", "Most Frequent Delay Location", "Most Frequent Bound"),
  align = 'c'
)
```

\clearpage

@fig-5 The summary table by month presents the mean delay times for each month
around 11.49 to 13.24, which fits the @fig-1 as most delay times are around 10 minutes,
that may slightly disrupt citizen's plan. However, some extreme cases up to 60 minutes
delay can totally destroy the citizen's reliability of public transit, TTC company
should try best to avoid the occurrence of extreme case. 

Based on table, there is a 
increase trend of delay frequency through the year, public transit system should
notice this situation and control the possibility of delay. West bound is the most
serve direction of delay problem, TTC could consider optimize the route or timetable
of the streetcar to reduce the possible delay. For most frequent delay's location, 
TTC should rewrite the extimate time table based on actual situation, in order to
make the estimate arrival time more exact, and increase the reliability of public transit.



\newpage

# References
